---
title: "0_descriptive_analysis"
output: html_document
---

```{r}

knitr::opts_chunk$set(echo = TRUE)

# load libraries
library(tidyverse)
library(readr)
# library(lme4)
# library(lmerTest)
# library(performance)

# import data 
self_responses_ids <- read_csv("data/self_responses_ids.csv")
```

We started by comparing the mean verbosity (word count per disclosure) between groups to see whether semantic structural differences might be driven by response length. Children ($M = 7.97$, $SD = 4.61$) and adults ($M = 6.811$, $SD = 6.25$) did not differ significantly in their average verbosity, $t(235.57) = 1.04, p = .298$. A 10,000-iteration bootstrap confirmed this null result, with the child mean falling well within the adult null distribution ($p_{emp} = .587$).


```{r verbosity test}

t.test(self_responses_ids %>% filter(age_group == "adult", is_skip == "FALSE") %>% select(word_count), 
       self_responses_ids %>% filter(age_group == "child", is_skip == "FALSE") %>% select(word_count))

```

There was no significant difference in Children's ($M = 7.97$, $SD = 4.61$) and Adults' ($M = 6.811$, $SD = 6.25$) verbosity,
$t(235.57) = 1.04, p = .298$.


```{r}

verbosity_df <- self_responses_ids %>%
  filter(!is_skip) %>%
  group_by(subject_id, age_group) %>%
  summarise(mean_word_count = mean(word_count), .groups = "drop")


child_verbosity_obs <- verbosity_df %>% 
  filter(age_group == "child") %>% 
  pull(mean_word_count) %>% 
  mean()

set.seed(420376)

boot_adult_verbosity <- replicate(10000, {
  verbosity_df %>%
    filter(age_group == "adult") %>%
    sample_n(18) %>%
    pull(mean_word_count) %>%
    mean()
})


p_val_verb <- sum(boot_adult_verbosity >= child_verbosity_obs) / 10000

p_val_verb
```

```{r}

verbosity_df %>%
  group_by(age_group) %>%
  summarise(
    mean_wc = mean(mean_word_count),
    sd_wc = sd(mean_word_count),
    n = n())

```

```{r}


```

