---
title: "sequential_analysis"
output: html_document
---
### Sequential Distance Analysis 

In this analysis, we calculated the Sequential Distance between responses.<br> 
**Formula: cosine distance = 1 — cosine similarity**


```{r}
knitr::opts_chunk$set(echo = TRUE)

# load libraries
library(tidyverse)
library(readr)

# import data 
# self_vectors_all <- read_csv("data/self_vectors_all.csv")
self_responses_ids <- read_csv("data/self_responses_ids.csv")
```

```{r sequential distances within participants}

# subtract the cosine similarity (between each response and the one immediately preceding it) from 1.

sequential_metrics <- self_responses_ids %>%
  group_by(subject_id) %>% # within participant
  filter(!is_skip) %>% # exclude skipped responses
  arrange(utterance) %>%  # capture temporal order of responses
  mutate(across(starts_with("Dim"), # shift vectors down by 1 row within each person
                ~ lag(.), 
                .names = "prev_{.col}")) %>%
  filter(!is.na(prev_Dim1_texts)) %>% # remove the first row per person (since it has no 'previous' response)
  rowwise() %>%
  mutate(
    
    # extract current vector (Dim1_texts to Dim768_texts)
    curr_vec = list(c_across(starts_with("Dim") & ends_with("_texts"))), 
    
    # extract previous vector (prev_Dim1_texts to prev_Dim768_texts)
    prev_vec = list(c_across(starts_with("prev_Dim") & ends_with("_texts"))),
    
    # calculate cosine distance: 1 - cosine similarity
    step_distance = 1 - (sum(curr_vec * prev_vec) / 
                         (sqrt(sum(curr_vec^2)) * sqrt(sum(prev_vec^2))))
  ) %>%
  ungroup()

```

```{r mean sequential distance per person}

# calculate mean distance per person, 1 value per participant

subject_sequential <- sequential_metrics %>%
  group_by(subject_id, age_group) %>%
  summarise(
    mean_step_distance = mean(step_distance, na.rm = TRUE),
    total_steps = n(),
    .groups = "drop"
  )

# subject_final <- subject_sequential %>% left_join(self_responses_ids, by = "subject_id")
```


```{r ANCOVA Age Group on Mean Step Distance}

# does age group influence the size of an individual's semantic leaps?
# df: subject_sequential -- contains (1) mean distance value per participant 

model_sequential <- lm(mean_step_distance ~ age_group, data = subject_sequential)
summary(model_sequential)

```

```{r}

trajectory_data <- sequential_metrics %>%
  group_by(subject_id, age_group) %>%
  summarise(
    mean_dist = mean(step_distance, na.rm = TRUE),
    se_dist = sd(step_distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(trajectory_data, aes(x = subject_id, y = mean_dist, color = age_group, group = age_group)) +
  geom_errorbar(aes(ymin = mean_dist - se_dist, ymax = mean_dist + se_dist), width = 0.2, alpha = 0.4) +
  geom_line(size = .6) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(values = c("adult" = "#2c7fb8", "child" = "#f03b20")) +
  labs(
    title = "Leaps in Self-Disclosure",
    subtitle = "Each dot represents an individual particiapant's mean semantic leap",
    x = "Participant",
    y = "Mean Cosine Distance (Semantic Leap)",
    color = "Age Group"
  ) +
  theme(legend.position = "bottom", text = element_text(size = 14),
        axis.text.x = element_blank())

```

To illustrate these group-level trends, we selected a representative child and adult whose semantic diversity scores were closest to their respective group medians. This ensures that the qualitative examples reflect the central tendency of each age group rather than statistical outliers.

```{r find representative participants}

example_participants <- subject_sequential %>%
  group_by(age_group) %>%
  # find the median out of all participant's mean distance, per group
  mutate(median_val = median(mean_step_distance)) %>%
  # difference between each participant's mean and the group median 
  mutate(dist_to_median = abs(mean_step_distance - median_val)) %>% 
  # select the participant with the smallest distance
  slice_min(dist_to_median, n = 3, with_ties = FALSE) %>%
  select(subject_id, age_group, dist_to_median, mean_step_distance, median_val, total_steps)

print(example_participants)

```

Finally, we graphed the distances in each step that our two exemplar participants took across all of their respective 12 responses.

```{r}

representative_steps <- sequential_metrics %>%
  filter(subject_id %in% c(child_id, adult_id))

ggplot(representative_steps, aes(x = utterance, y = step_distance, group = subject_id)) +
  # Line color locked
  geom_line(aes(color = age_group), size = 1.2) +
  # Point fill locked, with black border
  geom_point(aes(fill = age_group), size = 3, shape = 21, color = "black", stroke = 1) +
  # Explicitly naming the groups here prevents flipping
  scale_color_manual(values = c("adult" = "#2c7fb8", "child" = "#f03b20")) +
  scale_fill_manual(values = c("adult" = "#2c7fb8", "child" = "#f03b20")) +
  theme_minimal() +
  scale_x_continuous(breaks = 2:12) +
  labs(title = "Individual Comparison: Sequential Semantic Leaps",
       x = "Transition (Distance to Fact #)",
       y = "Cosine Distance",
       color = "Age Group",
       fill = "Age Group") +
  theme(text = element_text(size = 14))

```


